###Exercise 1 on ggplot 
1) How many rows are in penguins? How many columns?= 344 rows, 8 columns 

2) What does the bill_depth_mm variable in the penguins data frame describe? Read the help for ?penguins to find out = a number denoting bill length mm

3) Make a scatterplot of bill_depth_mm vs. bill_length_mm. That is, make a scatterplot with bill_depth_mm on the y-axis and bill_length_mm on the x-axis. Describe the relationship between these two variables:
```{r}
ggplot( data = penguins,
  mapping=aes(x=bill_length_mm, y=bill_depth_mm)
 ) +
  geom_point(aes(color=species,shape=species)) +
 geom_smooth(method="lm") +
  labs(title= "Bill length vs Bill depth", x="Bill length (mm)", y="Bill depth (mm)", color= "species", shape= "species")
```
In general, as bill length increases, bill depth decreases (negative association)

4) What happens if you make a scatterplot of species vs. bill_depth_mm? What might be a better choice of geom?
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=species, y=bill_depth_mm)
) +
 geom_point(aes(color=species,shape=species)) +
  geom_smooth(method="lm") +
 labs(title= "species vs bill depth", x="Species", y="Bill depth (mm)", color= "species", shape= "species")
```
A bar chart would be more suitable

5)Why does the following give an error and how would you fix it?= x and y values are missing, there is currently nothing to plot:
```{r}
ggplot(data = penguins) + 
  geom_point()
```

6) What does the na.rm argument do in geom_point()? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to TRUE? = n.arm, if FALSE, removed missing values with a warning. If TRUE, missing values silently removed. FALSE is the default:
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=bill_depth_mm, y=bill_length_mm)
) +
 geom_point(aes(color=species,shape=species), na.rm = TRUE) +
  geom_smooth(method="lm") +
 labs(title= "bill depth vs bill length", x="bill depth (mm)", y="Bill length (mm)", color= "species", shape= "species")
```

7) Add the following caption to the plot you made in the previous exercise: “Data come from the palmerpenguins package.” Hint: Take a look at the documentation for labs():
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=bill_depth_mm, y=bill_length_mm)
) +
 geom_point(aes(color=species,shape=species), na.rm = TRUE) +
  geom_smooth(method="lm") +
 labs(title= "bill depth vs bill length", x="bill depth (mm)", y="Bill length (mm)", color= "species", shape= "species", caption="data comes from the palmerpenguins package")
  
```
8) Recreate the following visualization. What aesthetic should bill_depth_mm be mapped to? And should it be mapped at the global level or at the geom level?:
```{r}
ggplot (
  data=penguins,
  mapping= aes(x=flipper_length_mm, y= body_mass_g)
  ) + 
  geom_point(aes(color=bill_depth_mm)) +
  geom_smooth() 
```

9) Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions:
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```

10) Will these two graphs look different? Why/why not?= No the maps would not look different as they are plotting the exact same thing
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )

```

#### ggplot2 calls
First 2 arguments for ggplot are data and mapping- don't need to type them out all the time- reduces amount of text and so makes it easier to see whats different between plots. A more concise way of writing the previous plot would be: 
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```
to create the same thing using a pipe (|>):
```{r}
penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```


#### Visualising Distributions
Visualisation depends on the variable type- for categorical, something like a bar chart is more suitable. The height of the bars displays how many observations occurred with each x value.
```{r}
ggplot(penguins, aes(x=species)) + geom_bar()
```
When levels unordered, preferable to order the bars based on frequency via transforming a variable into a factor addn reordering aforementioned factor:
```{r}
ggplot(penguins, aes(x=fct_infreq(species))) + geom_bar()
```
Quantitative data can be discrete or continuous. A histogram is a commonly used visualisation for continuous variables:
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```
Histograms divide the x axis into equally spaced bins- uses bar height to display no. of observations falling within each bin. Abovem tallest bar shows 39 obsv. have body_mass_g between 3500-3700g (left and right edges of bar)
Can set interval width via binwidth argument- measured in x variable units- different binwidth= different patterns
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 20)
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 2000)
```
Alternatively, density plots are also used to visualise numerical values. This is a smoothes out version of a histogram- good for continuous data with underlying smooth distribution. Shows fewer details than a histogram but easier to quickly see shape of distribution esp. when looking at skewness and modes
```{r}
ggplot(penguins, aes(body_mass_g)) + 
  geom_density()
```
### Exercises for Visualising Distributions
1) Make a bar plot of species of penguins, where you assign species to the y aesthetic. How is this plot different?= The plot is read in terms of length rather than height 
```{r}
ggplot(penguins, aes(y=species)) + geom_bar()
```

2) How are the following two plots different? Which aesthetic, color or fill, is more useful for changing the color of bars?= The first graoh only colours the outline of the bar whereas the second fills the bar with colour. Fill is better. 
```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```

3)What does the bins argument in geom_histogram() do?= States number of binds but overridden by binwidth- defaults to 30

4) Make a histogram of the carat variable in the diamonds dataset that is available when you load the tidyverse package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?= bindwidths between 0.01-1 reveal more interesting patterns
```{r}
ggplot(diamonds, aes(x=carat)) +geom_histogram(binwidth=0.2
                                      )
```


#### Visualising Relationships 
To visualise relationship between numerical and categorical variable, can use side-by-side box plots. Boxplot= visual shorthand measuring for position (percentiles) that describe distribution- useful for identifying potential outliers.
consists of 1. Box indicatung IQR (line in middle=median)- helps discern if distribution is symmetric around median or skewed 2. visual points- dispkay observations that fall >1.5 times IQR from either edge of box- unusual so plotted individually 3. line- extends from each end of box and goes to furthest non-outlier point in distribution
```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```
Alternatively, density plots- linewidth customises line thickness:
```{r}
ggplot(penguins, aes(x = body_mass_g, color = species)) +
  geom_density(linewidth = 0.75)
```
Can map species to colour and fill aestehtics- use alpha aesthetic to add transparency to filled density curves. This aesthetic takes values between 0 (completely transparent) and 1 (completely opaque).
```{r}
ggplot(penguins, aes(x = body_mass_g, color = species, fill = species)) +
  geom_density(alpha = 0.5)
```
To visualise 2 categorical variables, can use stacked bar charts:
```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```
Above plot displays each species frequency with respect to islands. Shows there is an equal number of Adelies on each island but not a good sense of % balance
```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
```
Above plot= relative frequency plot made by setting position= "fill" in geom- not affected by unequal numbers of penguins across islands
These bar charts map the variable that will be separated into bars to the x aesthetic, and the variable that will change the colors inside the bars to the fill aesthetic.
For 2 numerical variables: scatterplot:
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

For 3+ variables, can map them to additional aesthetics:
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island))
```
BUT adding too many aesthetics makes things cluttered- can also split plot into facets (subplots that each displays 1 set of data- esp. useful for categorical variables). The first argument of facet_wrap() is a formula3, which created with ~ followed by a variable name. The variable that passed to facet_wrap() should be categorical:
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  facet_wrap(~island)
```
 ### Exercises for Visualising Relationships
1) The mpg data frame that is bundled with the ggplot2 package contains 234 observations collected by the US Environmental Protection Agency on 38 car models. Which variables in mpg are categorical? Which variables are numerical? (Hint: Type ?mpg to read the documentation for the dataset.) How can you see this information when you run mpg?= Categorical= manufacturer, model, trans,drv, fl,class. Numerical=displ, year, cyl,cty,hwy

2) Make a scatterplot of hwy vs. displ using the mpg data frame. Next, map a third, numerical variable to color, then size, then both color and size, then shape. How do these aesthetics behave differently for categorical vs. numerical variables?
```{r}
ggplot(mpg, aes(x=displ,y=hwy)) + geom_point(aes(color= year, size = cty, shape = trans))
```

3) In the scatterplot of hwy vs. displ, what happens if you map a third variable to linewidth?
```{r}
ggplot(mpg, aes(x=displ,y=hwy)) + geom_point 
```


4) What happens if you map the same variable to multiple aesthetics?
```{r}
ggplot(mpg, aes(x=displ, y=hwy)) + geom_point(aes(color=fl, shape=fl, size=fl))
```

5) Make a scatterplot of bill_depth_mm vs. bill_length_mm and color the points by species. What does adding coloring by species reveal about the relationship between these two variables? What about faceting by species?
```{r}
ggplot(penguins, mapping = aes(x=bill_depth_mm, y=bill_length_mm)) + geom_point(aes(color=species))
```
```{r}
ggplot(penguins, mapping = aes(x=bill_depth_mm, y=bill_length_mm)) + geom_point()+facet_wrap(~species)
```
By adding the species variable, it can be seen that members of the same species have a similar bill depth/length to one another and are generally clustered together. Gentoo penguins tend to have average/long but shallow bills, Adelie penguins have deeper bills but they are shorter in length, and Chinstrap penguins seem to have longer and deeper beaks. 

6) Why does the following yield two separate legends? How would you fix it to combine the two legends?- remove the labs argument 
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() +
  labs(color = "Species")
```
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() 
```

7) Create the two following stacked bar plots. Which question can you answer with the first one? Which question can you answer with the second one?
```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```
The first graph tells us which penguin species are present per island as well as their proportion to one another within each island. The second graph tells us what proportion of penguin species are found in each island . For example, in Graph 1, it can be seen that both Adelie and Chinstrap penguins are found on Dream island in approximately equal numbers. However Graph 2 shows that 100% of chinstrap penguins are found on this island compared to appromixmately 1/3 of adelie penguins 

## Saving Plots
Once plot has been made, can save it out of R as an image to use elsewhere via ggsave()
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
ggsave(filename = "penguin-plot.png")
```
Saves plot to working directory. If width and height not specified, will be taken from dimensions of current plotting device BUT for reproducible code, would need to specify them. Generally recommended to assemble final reports using Quarto

//Exercises for Saving Plots 
1) Run the following lines of code. Which of the two plots is saved as mpg-plot.png? Why?
```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave("mpg-plot.png")
```
The scatter plot is saved as the ggsave command was written directly after it.

2) What do you need to change in the code above to save the plot as a PDF instead of a PNG? How could you find out what types of image files would work in ggsave()?
```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave("mpg-plot2.pdf")
```
instead of .png, .pdf. 









 








