###Exercise 1 on ggplot 
1) How many rows are in penguins? How many columns?= 344 rows, 8 columns 

2) What does the bill_depth_mm variable in the penguins data frame describe? Read the help for ?penguins to find out = a number denoting bill length mm

3) Make a scatterplot of bill_depth_mm vs. bill_length_mm. That is, make a scatterplot with bill_depth_mm on the y-axis and bill_length_mm on the x-axis. Describe the relationship between these two variables:
```{r}
ggplot( data = penguins,
  mapping=aes(x=bill_length_mm, y=bill_depth_mm)
 ) +
  geom_point(aes(color=species,shape=species)) +
 geom_smooth(method="lm") +
  labs(title= "Bill length vs Bill depth", x="Bill length (mm)", y="Bill depth (mm)", color= "species", shape= "species")
```
In general, as bill length increases, bill depth decreases (negative association)

4) What happens if you make a scatterplot of species vs. bill_depth_mm? What might be a better choice of geom?
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=species, y=bill_depth_mm)
) +
 geom_point(aes(color=species,shape=species)) +
  geom_smooth(method="lm") +
 labs(title= "species vs bill depth", x="Species", y="Bill depth (mm)", color= "species", shape= "species")
```
A bar chart would be more suitable

5)Why does the following give an error and how would you fix it?= x and y values are missing, there is currently nothing to plot:
```{r}
ggplot(data = penguins) + 
  geom_point()
```

6) What does the na.rm argument do in geom_point()? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to TRUE? = n.arm, if FALSE, removed missing values with a warning. If TRUE, missing values silently removed. FALSE is the default:
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=bill_depth_mm, y=bill_length_mm)
) +
 geom_point(aes(color=species,shape=species), na.rm = TRUE) +
  geom_smooth(method="lm") +
 labs(title= "bill depth vs bill length", x="bill depth (mm)", y="Bill length (mm)", color= "species", shape= "species")
```

7) Add the following caption to the plot you made in the previous exercise: “Data come from the palmerpenguins package.” Hint: Take a look at the documentation for labs():
```{r}
ggplot(
  data=penguins,
  mapping=aes(x=bill_depth_mm, y=bill_length_mm)
) +
 geom_point(aes(color=species,shape=species), na.rm = TRUE) +
  geom_smooth(method="lm") +
 labs(title= "bill depth vs bill length", x="bill depth (mm)", y="Bill length (mm)", color= "species", shape= "species", caption="data comes from the palmerpenguins package")
  
```
8) Recreate the following visualization. What aesthetic should bill_depth_mm be mapped to? And should it be mapped at the global level or at the geom level?:
```{r}
ggplot (
  data=penguins,
  mapping= aes(x=flipper_length_mm, y= body_mass_g)
  ) + 
  geom_point(aes(color=bill_depth_mm)) +
  geom_smooth() 
```

9) Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions:
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```

10) Will these two graphs look different? Why/why not?= No the maps would not look different as they are plotting the exact same thing
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )

```

#### ggplot2 calls
First 2 arguments for ggplot are data and mapping- don't need to type them out all the time- reduces amount of text and so makes it easier to see whats different between plots. A more concise way of writing the previous plot would be: 
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```
to create the same thing using a pipe (|>):
```{r}
penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```


#### Visualising Distributions
Visualisation depends on the variable type- for categorical, something like a bar chart is more suitable. The height of the bars displays how many observations occurred with each x value.
```{r}
ggplot(penguins, aes(x=species)) + geom_bar()
```
When levels unordered, preferable to order the bars based on frequency via transforming a variable into a factor addn reordering aforementioned factor:
```{r}
ggplot(penguins, aes(x=fct_infreq(species))) + geom_bar()
```
Quantitative data can be discrete or continuous. A histogram is a commonly used visualisation for continuous variables:
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```
Histograms divide the x axis into equally spaced bins- uses bar height to display no. of observations falling within each bin. Abovem tallest bar shows 39 obsv. have body_mass_g between 3500-3700g (left and right edges of bar)
Can set interval width via binwidth argument- measured in x variable units- different binwidth= different patterns
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 20)
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 2000)
```
Alternatively, density plots are also used to visualise numerical values. This is a smoothes out version of a histogram- good for continuous data with underlying smooth distribution. Shows fewer details than a histogram but easier to quickly see shape of distribution esp. when looking at skewness and modes
```{r}
ggplot(penguins, aes(body_mass_g)) + 
  geom_density()
```
### Exercises for Visualising Distributions
1) Make a bar plot of species of penguins, where you assign species to the y aesthetic. How is this plot different?= The plot is read in terms of length rather than height 
```{r}
ggplot(penguins, aes(y=species)) + geom_bar()
```

2) How are the following two plots different? Which aesthetic, color or fill, is more useful for changing the color of bars?= The first graoh only colours the outline of the bar whereas the second fills the bar with colour. Fill is better. 
```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```

3)What does the bins argument in geom_histogram() do?= States number of binds but overridden by binwidth- defaults to 30

4) Make a histogram of the carat variable in the diamonds dataset that is available when you load the tidyverse package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?= bindwidths between 0.01-1 reveal more interesting patterns
```{r}
ggplot(diamonds, aes(x=carat)) +geom_histogram(binwidth=0.2
                                      )
```


#### Visualising Relationships 








 








