---
title: "chapter 2"
output: html_document
date: "2025-01-21"
---
#### Coding Basics
can use to do basic maths:
```{r}
1/200*30
sin (pi/2)
```
can create new objects w assignment order:
```{r}
x<- 3*4
```
x stored not printed- can see stored variables in environment tab

can combine multiple elements into vector:
```{r}
prime <- c(2,3,5,7,11,13)
```
basic maths in vector is applied to every element:
```{r}
prime*2
```
all assignment statements have same form= object_name<- value

#### Testing Out Basics
```{r}
x
```
```{r}
random<-2.5 
```

#### Calling Functions
R has collection of built in functions- function_name(argument1 = value1, argument2 = value2, ...)
```{r}
seq(from=1, to =10)
```
often omit the names of the first several arguments in function calls:
```{r}
seq(1,10)
```
```{r}
x<- "hello"
```


### Exercises
1) Why does this code not work?:
```{r}
my_variable <- 10
my_varıable
#> Error: object 'my_varıable' not found
```
Variable name misspelt

2) Tweak each of the following R commands so that they run correctly:
```{r}
libary(todyverse)

ggplot(dTA = mpg) + 
  geom_point(maping = aes(x = displ, y = hwy)) +
  geom_smooth(method = "lm)
```
```{r}
library(tidyverse)

```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + geom_smooth(method = "lm")
```

3) Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?= shortcut menu comes up, also on help tab





