---
title: "chapter12"
output: html_document
date: "2025-03-04"
---
#### Logical Vectors
### Prerequisites
```{r PREREQUISITES}
library(tidyverse)
library(nycflights13)

x <- c(1, 2, 3, 5, 7, 11, 13)
x * 2


df <- tibble(x)
df |> 
  mutate(y = x * 2)
```

### Comparisons
A very common way to create logical vector = numeric comparsion with <, <= , >, >=, !=,and == 
have currently mostly created logical variables transiently via filter() =  - computed, used, then thrown away
```{r EXAMPLE: FILTER FINDS ALL DAYTIME DEPARTURES THAT ARRIVE ROUGHLY ON TIME}
flights |> 
  filter(dep_time > 600 & dep_time < 2000 & abs(arr_delay) < 20)
```
Useful to know that this is a shortcut - can explicityl create unnderlying logical variables w mutate:
```{r EXAMPLE CONT. USING MUTATE}
flights |> 
  mutate(
    daytime = dep_time > 600 & dep_time < 2000,
    approx_ontime = abs(arr_delay) < 20,
    .keep = "used"
  ) 
```
particularly useful for more complicated logic - na,ing intermediate steps makes it easier to read code + check each step has been computed correctly
initial filter equivalent to:
```{r FILTER EQUIVALENT TO}
flights |> 
  mutate(
    daytime = dep_time > 600 & dep_time < 2000,
    approx_ontime = abs(arr_delay) < 20,
  ) |> 
  filter(daytime & approx_ontime)
```


## FLoating Point COmpariosn
Beware of using == w numbers eg) following looks like vector contains numbers 1 and 2 but when tested for equality, get FALSE:
```{r BEWARE ==}
x <- c(1 / 49 * 49, sqrt(2) ^ 2)
x

x == c(1, 2)
```
why? computers store numbers with fixed number of decimal places - no way to exactly represent 1/49 or sqrt(2) 
subsequnet computations will be slightly off
can see exact values by calling print() with digits argument:
```{r DIGITS}
print(x, digits = 16)
```
R defaults to rounding these numbers as they are really close
now can see why == is failing, how can we fix it?
1 way is to use dplyr::near() - ignores small differences:
```{r NEAR}
near(x, c(1, 2))
```

## Missing Values
represent unknowns - "contagious" - operations involving unknown values will also be unknown
most confusing result is :
```{r CONFUSING RESULT}
NA == NA
```
easier to understand why if we articifically supply more context:
```{r THEORY AS TO WHY IT FAILS}
# We don't know how old Mary is
age_mary <- NA

# We don't know how old John is
age_john <- NA

# Are Mary and John the same age?
age_mary == age_john
#> [1] NA
# We don't know!
```

SO IF WANT TO FIND ALL FLIGHTS WHERE DEP_TIME IS MISSING, FOLLOWING CODE DOESN'T WORK - dep_time == NA
NA yilded for every single row
filter automatically drops missing values
Instead, use new tool = is.na()

## is.na()
works with any type of vector - returns TRUE for missing values and FALSE for anything else:
```{r IS.NA}
is.na(c(TRUE, NA, FALSE))

is.na(c(1,NA,3))

is.na(c("a", NA, "b"))
```
can use this argument to find all rows with missing dep_time:
```{r USING TO FIND MISSING DEP_TIME}
flights |> 
  filter(is.na(dep_time))
```
can also be useful in arrange() - usually puts all missing values at end but can override default by first sorting by is.na:
```{r USE IN ARRANGE}
flights |> 
  filter(month == 1, day == 1) |> 
  arrange(dep_time)

#vs

flights |> 
  filter(month == 1, day == 1) |> 
  arrange(desc(is.na(dep_time)), dep_time)
```

## Exercises
1. How does dplyr::near() work? Type near to see the source code. Is sqrt(2)^2 near 2?
```
source code = function (x, y, tol = .Machine$double.eps^0.5) 
{
    abs(x - y) < tol
}
MEANING = .Machine$double.eps = smallest positive floating - point number such that 1 + eps ! = 1
          tol = .Machine$double.eps^0.5 = square root of this small value, usually around 1.49e-8
          function checks iof absolute difference between x and y is smaller than tol
          
near function used to compare 2 numeric values for near - equality, accounting for floating - point error precision errors
in R, floating - point arithmetic can result in small rounding errors - helps check if 2 numbers are approximately equal by tolerating epsilon (small margin of error)
```
```{r 1}
sqrt(2)^2 == 2
near(sqrt(2)^2,2) # Yes, near 2 when using near
```
2. Use mutate(), is.na(), and count() together to describe how the missing values in dep_time, sched_dep_time and dep_delay are connected.
```{r 2}
flights |> 
  mutate(
    dep_time_na = is.na(dep_time),
    sched_dep_time_na = is.na(sched_dep_time),
    dep_delay_na = is.na(dep_delay)
  ) |> 
  count(dep_time_na, sched_dep_time_na, dep_time_na)
```
```
dep_time_na = FALSE - departure time not missing (flight departed)
sched_dep_time_na = FALSE - scheduled deprarture time never missing
dep_delay_na = FALSE -  departure delay not missing as fligth actually departed

dep_time_na = TRUE - departure time missing
sched_dep_time_na = FALSE
dep_delay_na = TRUE - departure delay missing - no flight = no delay

All rows with dep_time missing also have dep_delay missing
scheduled departure time never missing
missing values in dep_time and dep_delay indicate cancelled flights 
```


### Boolean Algebra 
once there are multiple logical vectors, can combine together using this
& = and, | = or, ! = not, xor() = exclusive or (xor(x, y) is true if x is true, or y is true, but not both)
eg) df |> filter(!is.na(x)) finds all rows where x isn't missing 
df |>  filter(x < -10 | x > 0) finds all rows where x is smaller than -10 or greater than 0
dont use && or || in dplyr functions - called short circuiting operators and only ever return single true or false - programming not data science

## Missing Values
rules for missing values in BA seem inconsistent at first:
```{r BOOLEAN NAs FIRST GLANCE}
df <- tibble(x = c(TRUE, FALSE, NA))
df |> 
  mutate(
    and = x & NA,
    oe = x | NA
  )
```
NA | TRUE - missing value in logical vector means that value could be either true or false
TRUE | TRUE / FALSE | TRUE are both TRUE because at least one of them is TRUE
NA | TRUE must also be ture because NA can be either true or false
BUT
NA | FALSE is NA because we don't know if NA is true or false - same reasoninh for NA & FALSE

## Order of Operations
doesnt work like in English:
```{r ALL FLIGHTS THAT DEPARTED IN NOVEMBER OR DECEMBER}
flights |> 
  filter(month == 11 | month == 12)
```
might be tempted to write code so as to say " find all flights that departed in november or december":
```{r INCORRECT VERSION OF ABOVE CODE}
flights |> 
  filter(month == 11 | 12)
```
no error message but doesnt seem to have worked
R evalutes month == 11 creating logical vectors (nov) - computes nov | 12
when use a number with logical operator, converts everything apart from 0 to TRUE - equivalent to nov | TRUE- will always be TRUE so every row selected:
```{r CORRECTION}
flights |> 
  mutate(
    nov = month == 11,
    final = nov | 12,
    .keep = "used"
  )
```

## %in%
easy way to  using == and | in right order is this 
x %in% y returns logical vector same length as x that is TRUE whenever value in x is anywhere in y:
```{r %IN%}
1:12 %in% c(1, 5, 11)

letters[1:10] %in% c("a", "e", "i", "o", "u")
```
so to find all flights in november and december:
```{r FLIGHTS IN NOV AND DEC USING %IN%}
flights |> 
  filter(month %in% c(11, 12))
```

%in% oberys different rules for NA to ==, as NA %in% NA is TRUE:
```{r %IN% HAS DIFFERENT RULES FOR NA}
c(1, 2, NA) == NA

c(1, 2, NA) %in% NA
```
can be a useful shortcut:
```{r USE AS A SHORTCUT}
flights |> 
  filter(dep_time %in% c(NA, 0800))
```

## Exercises
1. Find all flights where arr_delay is missing but dep_delay is not. Find all flights where neither arr_time nor sched_arr_time are missing, but arr_delay is.
```{r 1}
# flights where arr_delay missing but not dep_delay
missing_arr_delay <- flights |> 
  filter(is.na(arr_delay) & !dep_delay %in% NA)

# flights where neither arr_time or sched_arr_time are missing but arr_delay is 
missing_arr_delay_times <- flights |> 
  filter(!arr_time %in% NA & !sched_arr_time %in% NA & arr_delay %in% NA)

view(missing_arr_delay)
```

2. How many flights have a missing dep_time? What other variables are missing in these rows? What might these rows represent?
```{r 2}
# flights w missing dep_time
missing_dep_time <- flights |> 
  filter(is.na(dep_time))

#number of flights with missing dep_time
count_missing <- nrow(missing_dep_time)

#See which other variables are missing in these rows
missing_summary <- missing_dep_time |> 
  summarise(across(everything(), ~ sum(is.na(.)), .names = "missing_{.col}"))

missing_summary
```

3. Assuming that a missing dep_time implies that a flight is cancelled, look at the number of cancelled flights per day. Is there a pattern? Is there a connection between the proportion of cancelled flights and the average delay of non-cancelled flights
```{r 3}
#identify cancelled flights
cancelled <- flights |> 
  filter(is.na(dep_time))

#Number of cancelled a day
cancelled_daily <- cancelled |> 
  count(date = as.Date(time_hour)) |> 
  rename(cancelled_flights = n)

#calculate number of total flights per day 
total_flights <- flights |> 
  count(date = as.Date(time_hour)) |> 
  rename(total_flights = n)

#combine to get proportion of cancelled flights per day
cancelled_vs_total <- left_join(cancelled_daily, total_flights, by = "date") |> 
  mutate(proportion_cancelled = cancelled_flights / total_flights)

#calculate average delay of non-cancelled
avg_non_can_delay <- flights |> 
  filter(is.na(dep_time)) |> 
  group_by(date = as.Date(time_hour)) |> 
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE))

#jOIN DATASETES
final_data <- left_join(cancelled_vs_total, avg_non_can_delay, by = "date") 

# plot
ggplot(final_data, aes(x = proportion_cancelled, y = avg_delay)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs( title = "proportion of cancelled flights vs average delay of non cancelled flights",
        x = "proportion of cancelled flights", 
        y = "average delay of non-cancelled flights in mins") +
  theme_classic()
```

continue w plot