---
title: "chapter9"
output: html_document
date: "2025-01-29"
---
####### VISUALISE
#### Layers
### Aesthetic Mappings
mpg data frame bundled with the ggplot2 package contains 234 observations on 38 car models.
Among the variables in mpg are:

    displ: A car’s engine size, in liters. A numerical variable.

    hwy: A car’s fuel efficiency on the highway, in miles per gallon (mpg). A car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance. A numerical variable.

    class: Type of car. A categorical variable.
We can visualise the relationship between displ and hwy for various classes of cars. We can do this with a scatterplot where the numerical variables are mapped to the x and y aesthetics and the categorical variable is mapped to an aesthetic like color or shape:
```{r DISPL AND HWY BY CLASS}
# Left
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  scale_color_brewer(palette = "Set2")


# Right
ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
```
when class is mapped to shape, get 2 warnings:


    1: The shape palette can deal with a maximum of 6 discrete values because more than 6 becomes difficult to discriminate; you have 7. Consider specifying shapes manually if you must have them.

    2: Removed 62 rows containing missing values (geom_point()).
since ggplot only uses 6 shapes at a time, by default, additional groups will go unplotted when using shape aesthetic
second warning related -  62 SUVs in dataset - arent plotted

Similarly, we can map class to size or alpha aesthetics as well, which control the size and the transparency of the points, respectively:
```{r ALPHA FOR DISCRETE VARIABLE WARNING}
# Left
ggplot(mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point()


# Right
ggplot(mpg, aes(x = displ, y = hwy, alpha = class)) +
  geom_point()

```
Both mappings produce same warning - using alpha not suggested 
mapping an unordered discrete variable (class), to an ordered aesthetic (size/ alpha) generally not good - implies a ranking that doesnt exist 
After mapping aesthetic, ggplot selects a reasonable scale to use with the aesthetic, and it constructs a legend that explains the mapping between levels and values. 
For x and y aesthetics, ggplot2 does not create a legend, but it creates an axis line with tick marks and a label. 
The axis line provides the same information as a legend; it explains the mapping between locations and values
can also see visual properties of geom manually as argument of geom function (outside of aes()), instead of relying on variable mapping to determine appearances
eg) can make all points on  plot blue:
```{r BLUE}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(color = "blue")
```
Here, colour doesnt convey any informtion about variable- only changes plot appearance - need to pick a value that makes sense for that aesthetic:
    The name of a color as a character string, e.g., color = "blue"
    The size of a point in mm, e.g., size = 1
    The shape of a point as a number, e.g, shape = 1, 
    
### Exercises
1. Create a scatterplot of hwy vs. displ where the points are pink filled in triangles:
```{r 1}
ggplot(mpg, aes(x = hwy, y = displ)) +
  geom_point(colour = "pink", shape = "triangle")
```
2. Why did the following code not result in a plot with blue points?:
```{r 2}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, color = "blue"))
```
colours need to be set outside of the aesthetic mapping:
```{r 2}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(colour = "blue")
```
3. What does the stroke aesthetic do? What shapes does it work with?: Stroke controls the size of the edge/border of the points for filled circle, square, triangle, and diamond.
4. What happens if you map an aesthetic to something other than a variable name, like aes(color = displ < 5)? Note, you’ll also need to specify x and y.: 
```{r MAPPED BY OTHER THAN VARIABLE NAME}
ggplot(mpg, aes(x = displ, y = hwy, colour = displ < 5)) +
  geom_point()
```
creates logical variable with values TRUE AND FALSE for cars with displacement value below and above 5


### Geometric Objects
To change the geom in your plot, change the geom function that you add to ggplot():
```{r GEOMS}
# Left
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()

# Right
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_smooth()

```

Every geom function takes a mapping argument - either defined locally in geom layer or globally in ggplot() layer 
BUT not every aesthetic works with every geom - can set the shape of a point but cant set shape of line 
```{r GEOM_SMOOTH}
  # Left
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) + 
  geom_smooth()

# Right
ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) + 
  geom_smooth()
```


geom_smooth separates cars into 3 lines based on drv value -  describes car's drive train 
one line describes all the points that have an r value , one a f value and one a 4 value
can make it  clearer by overlaying lines of top of raw data and colour everything according to drv:
```{r COLOUR BY DRV}
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) + 
  geom_point() +
  geom_smooth(aes(linetype = drv))
```

plot contains 2 geoms in same graph
many geoms including geom_smooth use single geometric object to display multiple rows of data 
for these, can set group aesthetic to categorical variable to draw multiple objects 
ggplot2 will draw separate object for each unique value of the grouping variable
In practice, ggplot2 will automatically group the data for these geoms whenever you map an aesthetic to a discrete variable (as in the linetype example). It is convenient to rely on this feature because the group aesthetic by itself does not add a legend or distinguishing features to the geoms:
```{r GROUPING}
# Left
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth()

# Middle
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(group = drv))

# Right
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv), show.legend = FALSE)
```
If place mappings in geom function, ggplot treats them like local mappings for the layer- will use these mappings to extend/overwrite global mappings for that layer only
makes it possible to display different aesthetics in different layers:
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(aes(color = class)) + 
  geom_smooth()
```

