---
title: "chapter_4"
output: html_document
date: "2025-01-24"
---

#### Workflow: Code Style
```{r}
install.packages("styler")
```
access command palette via ctrl+shift+P and type styler
```{r}
library(tidyverse)
library(nycflights13)
```

### Names
As a general rule of thumb, it’s better to prefer long, descriptive names that are easy to understand rather than concise names that are fast to type
```{r}
# Strive for:
short_flights <- flights |> filter(air_time < 60)

# Avoid:
SHORTFLIGHTS <- flights |> filter(air_time < 60)
```


### Spaces
Put spaces on either side of mathematical operators apart from ^ (i.e. +, -, ==, <, …), and around the assignment operator (<-)
```{r}
# Strive for
z <- (a + b)^2 / d

# Avoid
z<-( a + b ) ^ 2/d
```
Don’t put spaces inside or outside parentheses for regular function calls. Always put a space after a comma
```{r}
# Strive for
mean(x, na.rm = TRUE)

# Avoid
mean (x ,na.rm=TRUE)
```
It’s OK to add extra spaces if it improves alignment. For example, if you’re creating multiple variables in mutate(), you might want to add spaces so that all the = line up.1 This makes it easier to skim the code
```{r}
flights |> 
  mutate(
    speed      = distance / air_time,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %%  100
  )
```


### Pipes
|> should always have a space before it and should typically be the last thing on a line. 
```{r}
# Strive for 
flights |>  
  filter(!is.na(arr_delay), !is.na(tailnum)) |> 
  count(dest)

# Avoid
flights|>filter(!is.na(arr_delay), !is.na(tailnum))|>count(dest)
```
If the function you’re piping into has named arguments (like mutate() or summarize()), put each argument on a new line. If the function doesn’t have named arguments (like select() or filter()), keep everything on one line unless it doesn’t fit, in which case you should put each argument on its own line.
```{r}
# Strive for
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights |>
  group_by(
    tailnum
  ) |> 
  summarize(delay = mean(arr_delay, na.rm = TRUE), n = n())
```
After the first step of the pipeline, indent each line by two spaces. RStudio will automatically put the spaces in for you after a line break following a |> . If you’re putting each argument on its own line, indent by an extra two spaces. Make sure ) is on its own line, and un-indented to match the horizontal position of the function name
```{r}
# Strive for 
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights|>
  group_by(tailnum) |> 
  summarize(
             delay = mean(arr_delay, na.rm = TRUE), 
             n = n()
           )

# Avoid
flights|>
  group_by(tailnum) |> 
  summarize(
  delay = mean(arr_delay, na.rm = TRUE), 
  n = n()
  )
```
OK to not do this if pipeline fits on 1 line:
```{r}
# This fits compactly on one line
df |> mutate(y = x + 1)

# While this takes up 4x as many lines, it's easily extended to 
# more variables and more steps in the future
df |> 
  mutate(
    y = x + 1
  )
```
Beware of pipelines >10-15 lines, try to break them up into smaller subtasks. Whenever there is a chance to give something an informative name, do it eg) fundamentally changing data structure- pivoting/summarising- means breaking up long pipelines if there are intermediate states that can get good names

### ggplot
Treat + the same way as |>
```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = month, y = delay)) +
  geom_point() + 
  geom_line()
```
if you can’t fit all of the arguments to a function on to a single line, put each argument on its own line
```{r}
flights |> 
  group_by(dest) |> 
  summarize(
    distance = mean(distance),
    speed = mean(distance / air_time, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = distance, y = speed)) +
  geom_smooth(
    method = "loess",
    span = 0.5,
    se = FALSE, 
    color = "white", 
    linewidth = 4
  ) +
  geom_point()
```


### Sectioning Comments
As script gets longer, utilise sectioning comments:
```{r}
# Load data --------------------------------------

# Plot data --------------------------------------
```
Keyboard shortcut to create these headers= ctrl+shift+R- will display them in the code navigation drop-down at the bottom-left of the editor


### Exercises
1) Restyle the following pipelines following the guidelines above:
```{r}
flights|>filter(dest=="IAH")|>group_by(year,month,day)|>summarize(n=n(),
delay=mean(arr_delay,na.rm=TRUE))|>filter(n>10)

flights|>filter(carrier=="UA",dest%in%c("IAH","HOU"),sched_dep_time>
0900,sched_arr_time<2000)|>group_by(flight)|>summarize(delay=mean(
arr_delay,na.rm=TRUE),cancelled=sum(is.na(arr_delay)),n=n())|>filter(n>10)
```

```{r}
flights |> 
  filter(dest == "IAH") |> 
  group_by(year, month, day) |> 
  summarise(
      n = n()),
      delay = mean(arr_delay, na.rm = TRUE)) |> 
  filter(n > 10)

flights |> 
  filter(
    carrier == "UA",
    dest %in% c("IAH", "HOU"),
    sched_dep_time > 0900,
    sched_arr_time < 2000) |> 
  group_by(flight) |> 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    cancelled = sum(is.na(arr_delay)), n = n()) |> 
  filter(n >10)
```



